# Reproducible Research: Peer Assessment 1
==========================================
Author: Nahir Rodriguez 

Date: August 17, 2014

---
```{r setoptions,echo = FALSE}
opts_chunk$set(echo = TRUE, results = "asis")
```

## Loading and preprocessing the data
Unzip and read the data.

```{r load data}
setwd("C:/Users/Nahir/Documents/RepData_PeerAssessment1")

unzip("activity.zip")

rawdb <- read.table("activity.csv",
                    header = T,sep = ",",na.strings = c("NA"),
                    colClasses = c("numeric","character","numeric"))

db <- rawdb

```

## What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.

```{r}
dbbydate <- aggregate(steps ~ date, data = db, FUN = "sum")
hist(dbbydate$steps,
     main = "Number of Steps per Day",
     xlab = "Steps")

meansteps <- mean(dbbydate$steps, na.rm = FALSE)
print(meansteps)
mediansteps <- median(dbbydate$steps, na.rm = FALSE)
print(mediansteps)
```

The mean total number of steps taken per day is `r format(meansteps,nsmall=0)`, while the median is `r format(mediansteps)`.


## What is the average daily activity pattern?
Make a time series plot  of the 5-minute interval and the average number of steps taken, averaged across all days

```{r}
dbdactivity <- aggregate(steps ~ interval, data = db, FUN = "mean")

plot(dbdactivity$interval, dbdactivity$steps,type = "l",
     main = "Activity per 5-Minute Interval",
     xlab = "5-Minute Interval",
     ylab = "Steps"
     )

maxsteps <- max(dbdactivity$steps)
print(maxsteps)
maxstepsinterval <- dbdactivity[which.max(dbdactivity$steps),c("interval")]
print(maxstepsinterval)
```

The 5-minute interval, which, on average across all the days in the dataset, contaied the maximum number of steps was `r maxstepsinterval` with `r format(maxsteps,nsmall=1)` steps.

## Imputing missing values


```{r}
nadb <- rawdb[is.na(rawdb$steps) | 
                  is.na(rawdb$date) | 
                  is.na(rawdb$interval),]
narows <- nrow(nadb)
narows
```

The total number of missing values in the dataset (i.e. the total number of rows with NAs) is `r narows`. To clean the data the corresponding mean of each 5-minute interval was used to replace NA values.


```{r}
completedb <- rawdb[!is.na(rawdb$steps) & 
                        !is.na(rawdb$date) & 
                        !is.na(rawdb$interval),]

nadb <- nadb[,c("date","interval")]
nadb <- merge(nadb,dbdactivity,by="interval")

completedb2 <- rbind(completedb,nadb)

cleandb <- merge(rawdb[,c(c("date","interval"))],completedb2,by=c("date","interval"),sort = FALSE)

nrow(cleandb)
nrow(nadb)+nrow(completedb)

cleandbbydate <- aggregate(steps ~ date, data = cleandb, FUN = "sum")

hist(cleandbbydate$steps,
     main = "Number of Steps by Day (Clean Data Base)",
     xlab = "Steps")

cleanmeansteps <- mean(cleandbbydate$steps, na.rm = FALSE)
print(cleanmeansteps)
cleanmediansteps <- median(cleandbbydate$steps, na.rm = FALSE)
print(cleanmediansteps)

```

Note that using the mean for that 5-minute interval to repalce NA values, the mean total number of steps taken per day is `r format(cleanmeansteps,nsmall=2)`, while the median is `r format(cleanmediansteps,nsmall=2)`. Very similar to the results with NA values included.



## Are there differences in activity patterns between weekdays and weekends?

```{r}
cleandb$day <- weekdays(as.Date(cleandbbydate$date,"%Y-%m-%d"))

cleandb$daytype <- ""

cleandb[cleandb$day %in% c("Monday","Tuesday","Wednesday","Thursday","Friday"),c("daytype")] <- "weekday"
cleandb[cleandb$day %in% c("Saturday","Sunday"),c("daytype")] <- "weekend"

dbdactivity <- aggregate(steps ~ interval + daytype, data = cleandb, FUN = "mean")


library(lattice)
xyplot(steps ~ interval | daytype, 
       data = dbdactivity, 
       layout = c(1, 2),
       type="a",
       xlab = "Interval",
       ylab = "Number of steps")

```

There are differences in activity patterns between weekdays and weekends as seen in the graph above.
